Assessing profiles of various loan borrowers by Varsha Shewale
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(reshape2)
library(dplyr)
library(GGally)
library(Hmisc)
library(corrplot)
library(PerformanceAnalytics)
library(psych)
library(memisc)
library(ordinal) 
library(MASS)
library(nnet)

```

```{r echo=FALSE, warning=FALSE,Load_the_Data}
# Load the Data
path <- 
  paste("/Users/varsha/Documents/Udacity_DAND_t1t2/Term2/ProjectSubmissions",
             "/P2_R_EDA/prosperLoanData.csv",sep ="")
loan_whole<-read.csv(path)

```

The original dataset has about 81 variables. Out of these I am analyzing only 
the below mentioned 12(now 13) variables which I am more interested in viz.
"ListingNumber","Term","LoanStatus",      "Occupation","EmploymentStatus","CreditScoreRangeLower","CreditScoreRangeUpper",
"DelinquenciesLast7Years","StatedMonthlyIncome","DebtToIncomeRatio",
"IncomeRange","LoanOriginalAmount","ListingCategory"

```{r echo=FALSE, warning=FALSE, Subset_data_with_necessary_variables}
loan<-subset(loan_whole, select = c("ListingNumber","Term","LoanStatus",      "Occupation","EmploymentStatus","CreditScoreRangeLower","CreditScoreRangeUpper",
"DelinquenciesLast7Years","StatedMonthlyIncome","DebtToIncomeRatio",
"IncomeRange","LoanOriginalAmount","ListingCategory..numeric."))

#changing column name for ease and better readability
names(loan)<-c("ListingNumber","Term","LoanStatus",      "Occupation","EmploymentStatus","CreditScoreRangeLower","CreditScoreRangeUpper",
"DelinquenciesLast7Years","StatedMonthlyIncome","DebtToIncomeRatio",
"IncomeRange","LoanOriginalAmount","ListingCategory")

```

Further I will be decoding the values in the ListingCategory with category names 
Reference:
[https://stackoverflow.com/questions/25432935/how-to-replace-values-with-table-map-in-r](replace-values-with-table-map-in-r)

```{r echo=FALSE, warning=FALSE, Decoding_ListingCategory}
#id <- sapply(0:20,function(x)toString(x))
id<-c(0:20)
name<-c("Not Available", "Debt Consolidation","Home Improvement", 
"Business", "Personal Loan", "Student Use", "Auto"," Other","Baby&Adoption", 
"Boat", "Cosmetic Procedure","Engagement Ring","Green Loans", 
"Household Expenses","Large Purchases", "Medical/Dental","Motorcycle","RV", 
"Taxes", "Vacation","Wedding Loans")

map_listingCategory<-data.frame(id,name)
loan$ListingCategory <- with(map_listingCategory, 
                             name[match(loan$ListingCategory, id)])


```

The subsetted dataset has 113,937 observations of 12 variables (now 13).

# Univariate Analysis

Below is the summary of the loan data subset
```{r echo=FALSE,warning=FALSE, Summary}
summary(loan)
```

Further I wish to analyze LoanStatus.
Looking at the LoanStatus as a percentage plot

```{r echo=FALSE, warning=FALSE, Univariate_Plots}
#Analyzing LoanStatus
ggplot(data = loan, aes(x = LoanStatus))+
    geom_histogram(stat=  'count',aes(y=100*(..count..)/sum(..count..)))+
    labs(y ="Percentage")+ 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Looking at the LoanStatus using a log scale

```{r echo=FALSE,warning=FALSE}
ggplot(data = loan, aes(x = LoanStatus))+
    geom_histogram(stat=  'count')+
    scale_y_log10(breaks = c(0,100,250,500,1000,5000,20000,40000,60000))+
    labs(y ="Count (log scale")+ 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Around 10% were charged-off, 5% defaulted on their loan, approx. 2% are past 
their due payment date. 
Around 33% have completed paying off their loan. Around 50% have 'current' 
loan status.

Let's look at the loan term.

```{r echo=FALSE,warning=FALSE}
ggplot(data = loan, aes(Term))+geom_bar(aes(y=100*(..count..)/sum(..count..)))
```

Approx. 21% borrowers have a loan term of 60 months, 
78% borrowers have a term of 36 months; and 
approx. 2% have a term of 12 months.

Let's look at the Employment Status with percent defaulting/chargedoff.

```{r echo=FALSE, warning = FALSE}
#Analyzing Employment Status
ggplot(data = loan, aes(EmploymentStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..),
               fill = LoanStatus %in% c('Defaulted','Chargedoff') ))+
  labs(y="Employment Status as a Percentage(%)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
```

60% of Loan borrowers are Employed, 24%(approx.) are fulltime employees,approx. 
5% are self employed, about 1% are unemployed, approx. 5% do not have any
employment status available( This is definitely not good as employment status is
key in determining whether the borrower is capable of paying off his loan)
Right now it is not clear if a specific employment category is 
defaulting/chargedoff more than the others.

Let's look at the Credit Scores.
For Analyzing credit scores, I have created a CreditScoreMean column which is 
average of CreditScoreRangeLower and CreditScoreRangeUpper.

```{r echo=FALSE, warning = FALSE}
#Analyzing credit score ranges
loan$CreditScoreMean<-round(rowMeans(subset(loan, 
      select = c(CreditScoreRangeLower, CreditScoreRangeUpper)), na.rm = TRUE))

summary(loan$CreditScoreMean)
ggplot(data = loan, aes(CreditScoreMean))+geom_histogram(binwidth = 70)+
  scale_x_continuous(breaks = seq(0,1000,100))
```

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan, aes(x = "", y = CreditScoreMean))+geom_boxplot()

```

CreditScoreMean has an IQR = 60, 50% of the borrowers have CreditScoreMean 
between 670 and 730, maximum CreditScoreMean is 890, with minimum being at 10

Let's look at the delinquencies for last 7 years.

```{r echo=FALSE, warning = FALSE}
summary(loan$DelinquenciesLast7Years)
ggplot(data = loan, aes(DelinquenciesLast7Years))+geom_histogram(binwidth = 2)+
  scale_x_continuous(breaks = seq(0,110,5))

ggplot(data = loan, aes(DelinquenciesLast7Years))+
  geom_histogram(binwidth = 10,alpha = 0.5)+
  scale_y_log10(breaks = c(0,1,2,5,10,50,1000,10000,50000,80000))+
  scale_x_continuous(breaks = seq(0,100,10))

ggplot(data = loan, aes(x = "", y = DelinquenciesLast7Years))+geom_boxplot()
```

Around 80000 (70%) borrowers have zero delinquencies in the past 7 years.
Further subsetting borrowers with one or more delinquencies.

```{r echo=FALSE, warning = FALSE}
loan.deliquent<-subset(loan,loan$DelinquenciesLast7Years>=1)
```

Approx. 32% borrowers were delinquent in the past 7 years.

```{r echo=FALSE, warning = FALSE}
#Analyzing stated monthly income
summary(loan$StatedMonthlyIncome)

```

Considering income, mean income is $5608 and median income is $4667. 
Some outliers are driving the mean higher.

```{r echo=FALSE, warning = FALSE}

#Using boxplot to see the distribution and outliers if any
ggplot(data = loan, aes(x="", y=StatedMonthlyIncome))+geom_boxplot(alpha = 0.1)

#Above plot squisshes the boxplot's IQR due to outliers
ggplot(data = loan, aes(x="", y=StatedMonthlyIncome))+ 
  geom_boxplot()+
  coord_cartesian(ylim = c(0,10000))
#75% of borrowers have monthly income <=$6825

ggplot(aes(x=StatedMonthlyIncome), data = loan)+geom_histogram(binwidth = 1000)+
  scale_x_continuous(limits = c(0,25000))

```

As anticipated the income distribution is right skewed following the 80-20 rule
with very few people having very high incomes and vice versa.

Let's look at LoanOriginalAmount.

```{r echo=FALSE, warning = FALSE}
summary(loan$LoanOriginalAmount)
ggplot(aes(x=LoanOriginalAmount), data = loan)+
  geom_histogram(binwidth = 5000)+
  scale_x_continuous(breaks = seq(0,30000,5000))

ggplot(aes(x ="", y=LoanOriginalAmount), data = loan)+geom_boxplot()+
  scale_y_continuous(breaks = seq(0,30000,5000))
```

Considering the loan amounts borrowed, based on the boxplot 75% of the borrowers
have loan amount <= $12000 and the maximum loan amount is $35000.

```{r echo=FALSE, warning = FALSE}
sum(loan$LoanOriginalAmount > 12000)
```

Out of 113,937 borrowers, 26909 (25%) have loan amount > $12000 

##Summarizing findings of Univariate Analysis

1. Out of 113,937 borrowers, 26909 (25%) have loan amount > 12000, median
loan amount is equal to $6500, maximum loan amount is $35000 with minimum loan 
amount = $1000
2. 75% of borrowers have monthly income <=$6825 with maximum monthly income at
$1,750,000. This caused an extremely right-skewed distribution.
3. 60% of Loan borrowers are Employed, 24%(approx.) are fulltime employees,
approx. 5% are self employed, about 1% are unemployed, approx. 5% don't have any
employment status available( This is definitely not good as employment status is
key in determining whether the borrower is capable of paying off his loan)
4. Majority of the borrowers (approx. 78%)  have a term of 36 months 
5. 33% completed their loan, around 10% were charged-off, 5% defaulted
6. CreditScoreMean is a calculated column using the average of lowerCreditScore
range and upperCreditScore range for each borrower
CreditScoreMean has an IQR = 60, 50% of the borrowers have CreditScoreMean 
between 670 and 730, maximum CreditScoreMean is 890, with minimum being at 10 
(would definitely like to investigate the borrowers with lower CreditScoreMean)
7. Approx. 32% borrowers have 1 or more delinquencies in past 7 years.

Considering the above findings, I further wish to explore:
- borrowers with low credit scores and/or unemployed.
- profiles of people who were chargedoff and defaulted.
- profiles of people who completed their loans.
- if past deliquencies have any effect on today's loan status.

#Bivariate Analysis (part 1)

NOTE: 
BASED ON SUBSETS DRAWN IN FURTHER ANALYSIS, SOMETIMES I FELT THE NEED TO
CONDUCT UNIVARIATE ANALYSIS TO LOOK AT THE DISTRIBUTION OF A KEY VARIABLE WITHIN  
THE SUBSET ALONE.

Let's analyze creditScoreMean vs. loan Amount

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan, aes(x = CreditScoreMean, y = LoanOriginalAmount))+
  geom_point()
```

Loan Amount vs. CreditScoreMean yeilds a left skewed distribution. We can see 
that with an increase in the creditScoreMean borrowers are able to avail
higher amount loans as expected. The surprising aspect of the plot is loans 
lent to borrowers with low credit scores (almost close to 10).

Let's find the coefficient of corelation between Loan Amount and CreditScoreMean

```{r echo=FALSE, warning = FALSE}
cor(loan$CreditScoreMean, loan$LoanOriginalAmount,use = 'complete.obs') 
```

The correlation coefficient is 0.3408, which means the relation is positive and 
meaningful but not very strong.

Lets analyze these low credit score borrowers.

```{r echo=FALSE, warning = FALSE}
loan.low_credit <-subset(loan,loan$CreditScoreMean <=10)
#Borrowers with credit score mean <= 10 are 133. Now lets see their loan status
ggplot(data = loan.low_credit,aes(x = LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))
```

Out of the 133 low credit score borrowers, approx.
- 29% completed their loan,
- 26% were chargedoff,
- 44% defaulted

Let's create same plot for high credit score borrowers

```{r echo=FALSE, warning = FALSE}
loan.high_credit <-subset(loan,loan$CreditScoreMean > 10)
#Borrowers with credit score mean > 10 are 133. Now lets see their loan status
ggplot(data = loan.high_credit,aes(x = LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

  
```

Out of 113213 high credit score borrowers, 
- 33% completed their loans,
- 10% were chargedoff,
- 4% defaulted,

Clearly low credit score is a key attribute in determining future loan payment 
status as the percentage of defaulters/chargedoff is much higher with low credit
score borrowers.

Let's subset the low credit score borrowers to keep only defaulted and 
chargedoff cases.

```{r echo=FALSE, warning = FALSE}
loan.low_credit_defChar<-subset(loan.low_credit,loan.low_credit$LoanStatus %in% 
                             c("Chargedoff","Defaulted"))
```
Lets explore their occupation and employment status.

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan.low_credit_defChar,aes(x = EmploymentStatus))+
  geom_bar()

#**************

```

Out of the 133 lower credit score cases,where 94 are defaulting or chargedoff 
cases, no data is available regarding their EmploymentStatus. It is either
stated as 'Not available' or is missing.
Let's check their occupation.

```{r echo=FALSE, warning = FALSE}
unique(loan.low_credit_defChar$Occupation)
ggplot(data = loan.low_credit_defChar, aes(x = Occupation))+
  geom_bar(stat = "count")+coord_flip()
```

The defaulters/chargeoff borrowers with lower credit score mean seem to have
myriad occupations.
Lets explore their monthly income

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan.low_credit_defChar, aes(StatedMonthlyIncome))+
  geom_histogram()

summary(loan.low_credit_defChar$StatedMonthlyIncome)
ggplot(data = loan.low_credit_defChar,aes(x = "",y = StatedMonthlyIncome))+
  geom_boxplot()+coord_cartesian(ylim=c(0,10000))
```

IQR indicates 50% of low credit defaulters/chargedoff have income in range $1200
to $2800 approx.
It is surprising to see someone with monthly income of $33330 with credit score
mean <= 10 and being in the defaulting/chargedoff category.

```{r echo=FALSE, warning = FALSE}
a<-subset(loan,StatedMonthlyIncome >=30000 & CreditScoreMean <=10 )
```

The above borrower had taken a loan =$3600 for 36 months term and he defaulted 
on his loan. There is no other data available for this specific defaulter.

Moving on to our other borrowers in the above subset "loan.low_credit_defChar"
and comparing them to the rest of the borrowers.

```{r echo=FALSE, warning = FALSE}
loan.low_credit_non_defChar<-subset(loan.low_credit,
                         !(LoanStatus %in% c("Defaulted","Chargedoff")) )
```

```{r echo=FALSE, warning = FALSE}
summary(loan.low_credit_non_defChar$StatedMonthlyIncome)
ggplot(data = loan.low_credit_non_defChar,aes(x = "",y = StatedMonthlyIncome))+
  geom_boxplot()
```

Lets do a side by side comparison of Incomes between low credit 
defaulters/Chargedoff and other low credit borrowers

```{r echo=FALSE, warning = FALSE}
temp_df <- merge(loan.low_credit_non_defChar$StatedMonthlyIncome,
                  loan.low_credit_defChar$StatedMonthlyIncome) 
names(temp_df)<-c("lc_non_dc","lc_dc")
t<-melt(temp_df)

ggplot(data = t, aes(x=variable, y= value))+geom_boxplot()+
  coord_cartesian(ylim = c(0,10000))
```

As seen from the side by side boxplots, the income distribution between low 
credit defaulters/chargedoff and low credit others is NOT much different. 
Infact the numbers seen in summary stats for low credit non defaulters non 
chargedoff are lower looking at the IQR and mean!

#Bivariate Analysis (part 2)

Now let's explore relation between creditScoreMean and deliquencies

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan, aes(y = CreditScoreMean,x=DelinquenciesLast7Years))+
  geom_point(alpha = 0.7)

```

The graph of the CreditScore Vs. Deliquencies does not show a straight-forward
relation. I was hoping to see a negative corelation as higher number of 
deliquencies should adversely affect the credit mean but even with as high as 
100 deliquencies the credit score mean is in 500-750 range approx.

Let's explore StatedMonthlyIncome vs. deliquencies

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan, aes(y = StatedMonthlyIncome,x=DelinquenciesLast7Years))+
  geom_point(alpha = 0.3)+
  scale_y_log10(breaks = c(0,1000,2000,3000,4000,5000,6000,7000,10000))+
  coord_cartesian(ylim = c(1000,10000))


ggplot(data = loan, aes(y = StatedMonthlyIncome,x=DelinquenciesLast7Years))+
  geom_point(alpha = 0.3)+
  coord_cartesian(ylim = c(1000,10000))

```

This graph is also same as above, with no straight-forward relation between 
income and deliquencies. Borrowers with small as well as large incomes have
varying numbers of deliquencies.

Now let's look into delinquencies vs. debt-to-income ratio

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan, aes(y = DebtToIncomeRatio,x=DelinquenciesLast7Years))+
  geom_point(alpha = 0.7)+
  scale_y_continuous(breaks = c(0,10,0.5))
```

In general there is negative corelation here, the lower the Debt-to-Income ratio
the higher are the deliquencies.
The number of past deliquencies seen is very high with debt-to-income ratio of 
around 0.5
The correlation coefficient is:
```{r echo=FALSE, warning = FALSE}
cor(loan$DelinquenciesLast7Years,loan$DebtToIncomeRatio, use = 'complete.obs')
```
-0.04387671.
So the correlation is negative but not meaningful as it's absolute value < 0.3.

Let's see if there is any correlation between loan amount and past deliquencies.

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan, aes(y = LoanOriginalAmount,x=DelinquenciesLast7Years))+
  geom_point(alpha = 0.7)
```

In general there is a negative corelation between loan amount and delinquencies.
Borrowers with large number of past delinquencies have taken smaller loan 
amounts.

The correlation coefficient is:
```{r echo=FALSE, warning = FALSE}
cor(loan$DelinquenciesLast7Years,loan$LoanOriginalAmount, use = 'complete.obs')
```
-0.1366667.
So the correlation is negative but not meaningful as it's absolute value < 0.3

#Part Bivariate Analysis (part 3)

Let's see if past delinquencies show any impact on CreditScoresMean.

```{r echo=FALSE, warning = FALSE}
ggplot(loan, aes(y = CreditScoreMean,x = DelinquenciesLast7Years))+
  geom_point()+
  scale_y_continuous(breaks = seq(0,900,100))

```

Let's explore past delinquencies versus loan status.

```{r echo=FALSE, warning = FALSE}

ggplot(loan, aes(y = LoanStatus,x = DelinquenciesLast7Years))+geom_point()

```

Looking at the past delinquencies its difficult to predict future loan's payment 
status.
I was hoping to see fewer past deliquencies for Loan Status = 'Completed'.

```{r echo=FALSE, warning = FALSE}
ggplot(loan.deliquent, aes(x= LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  scale_y_continuous(breaks = seq(0,60,10))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Among borrowers with past delinquencies, 
-around 14% were chargedoff,
-around 5% defaulted,
-around 33% completed their loan.

Lets try to see same plot for borrowers with zero past deliquencies

```{r echo=FALSE, warning = FALSE}
loan.nondelinquent<-subset(loan,loan$DelinquenciesLast7Years <= 0)

ggplot(loan.nondelinquent, aes(x= LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  scale_y_continuous(breaks = seq(0,60,10))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

67% of borrowers have zero past delinquencies (for last 7 yrs).

Among borrowers with zero past delinquencies, 
-around 9% were chargedoff,
-around 4% defaulted,
-around 33% completed their loan.

Comparing borrowers with past deliquencies vs those with zero deliquencies, we 
see that ones with past delinquencies have higher percentages of chargedoff/
defaulting loans.

#Bivariate Analysis (part 4)

Now let's look at people who availed loan and are unemployed.

```{r echo=FALSE, warning = FALSE}
#Analyzing Employment Status and Loan Status
ggplot(data = loan, aes(EmploymentStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..),
               fill = LoanStatus %in% c('Defaulted','Chargedoff') ))+
  labs(y="Employment Status as a Percentage(%)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Borrowers with all types of 'Employment Status' had either defaulted/chargedoff.
Want to see the profile of the unemployed borrowers 

```{r echo=FALSE, warning = FALSE}
loan.unemp <- subset(loan, EmploymentStatus == 'Not employed')
str(loan.unemp) #returns 835 borrowers (not employed)
```

```{r echo=FALSE, warning = FALSE}
ggplot(loan.unemp,aes(x=LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Among the unemployed borrowers, 
 23% are chargedoff loans,
 3% are defaulting loans,
 41% are completed loans

Let's compare above with employed borrowers (Employed, full-time, Other,
part-time, self employed)

```{r echo=FALSE, warning = FALSE}
loan.emp<-subset(loan, loan$EmploymentStatus %in% 
            c("Employed","Full-time","Other","Part-time","Self-employed"))

ggplot(loan.emp,aes(x=LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

Around 92% are employed borrowers based on our definition of 'employed' as given
in the vector above.
Among the employed borrowers, 
- 10% are chargedoff loans, 
- 3% are defaulting loans,
- 31% are completed loans

Comparing employed and unemployed borrowers, the percentage of loans getting 
chargedoff is very high with unemployed borrowers.

Lets look at occupations of unemployed borrowers.

```{r echo=FALSE, warning = FALSE}
unique(loan.unemp$Occupation) # returns all possible occupations.
#Lets see if any specific occupation is higher than the rest

ggplot(data = loan.unemp, aes( x=Occupation))+geom_bar()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Approx. 800 of 835 unemployed have listed 'other' as their occupation while
the majority of the remaining appear to be students.
This finding was something unexpected for me as I was anticipating the 
unemployed borrowers to be mostly students.
Below I am recreating above plot excluding 'Other' occupation.

```{r echo=FALSE, warning = FALSE}
ggplot(data = subset(loan.unemp, Occupation != 'Other'), aes( x=Occupation))+geom_bar()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Lets analyze the loan amounts that were taken by the unemployed borrowers

```{r echo=FALSE, warning = FALSE}
ggplot(aes(y = LoanOriginalAmount, x =""), data = loan.unemp) +
  geom_boxplot()

summary(loan.unemp$LoanOriginalAmount) 

```

50% of unemployed borrowers availed loan amounts between $2500 and $6000. 
Median being $4000 and maximum being $25000. 

Digressing a bit to see how much loan do students borrow in general.

```{r echo=FALSE, warning = FALSE}
loan.student<- loan[grep("Student", loan$Occupation), ]
```

Out of 113937 borrowers, 699 are students. Checking their employment status

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan.student, aes(x = EmploymentStatus))+ geom_bar()

ggplot(data = loan.student, aes(x = EmploymentStatus))+ 
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  ylab("Employment % of students in each category")

```

It was fascinating to find only about 3% students unemployed. Rest were either 
working part-time, full-time, self-employed, etc.

Lets look at the loan amount borrowed by students along with loan status

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan.student, aes(x = LoanOriginalAmount))+ 
  geom_histogram(binwidth = 2000)

ggplot(data = loan.student, aes(x ="",y = LoanOriginalAmount))+ 
  geom_boxplot()

summary(loan.student$LoanOriginalAmount)
```

Its a right skewed distribution. 50% of students borrowed loan between $2000 and
$5000, with maximum loan at $20000.
I am wondering if all these loans were for education or were personal, etc.
Need to see if the whole dataset had a Loan_type variable.
The dataset actually has a variable 'ListingCategory' which mentions type of 
loan. Adding this variable to our dataframe by changing code in the beginning.
Decoding the values in the ListingCategory with category names. Reference:
https://stackoverflow.com/questions/25432935/how-to-replace-values-with-table-
map-in-r
And then viewing the barchart for ListingCategory

```{r echo=FALSE, warning = FALSE}
#basic bar chart of ListingCategory

ggplot(data = loan,aes(x = ListingCategory))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#percentage bar plot
ggplot(data = loan,aes(x = ListingCategory))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

It is interesting to see that around 51% borrowers have borrowed money to pay 
off their other loans (based on ListingCategory : Debt Consolidation)
This makes me want to digress a bit more and analyze these individuals more
in terms of LoanStatus

```{r echo=FALSE, warning = FALSE}
loan.debtConsolidate<-loan[loan$ListingCategory == 'Debt Consolidation',]

#barchart for loanStatus
ggplot(data = loan.debtConsolidate,aes(x = LoanStatus))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Based on the above barchart, approx. 22% of debt consolidators have completed
their loan, 6% chargedoff, 2% defaulted, whereas about 67% have status listed as
'current' which makes me think if all Debt Consolidation related loans are 
pretty recent compared to other loans.

Moving my focus back to students...
Redrawing the ListingCategory barchart

```{r echo=FALSE, warning = FALSE}
#percentage bar plot
ggplot(data = loan,aes(x = ListingCategory))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = loan,aes(x = ListingCategory))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Based on the barchart around 1% (approx. 1000) of total borrowers have taken 
loan for student_use.
Now creating similar barchart just for subsetted student dataframe

```{r echo=FALSE, warning = FALSE}

ggplot(data = loan.student,aes(x = ListingCategory))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = loan.student,aes(x = ListingCategory))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

So out of approx. 1140 borrowers with occupation listed as students less than 
100 have ListingCategory as 'student_use'. Approx. 130 (18%) have listed 'Debt 
Consolidation' and for 270 students the data is unavailable.
It is surprising to see almost 18% students taking loans for paying off their 
old loans.

Moving back to unemployed borrowers... 
Filtering unemployed borrowers with low credit scores.

```{r echo=FALSE, warning = FALSE}

loan.unemp_lowCredit<-
  loan %>%
  filter(EmploymentStatus == "Not employed" & CreditScoreMean <11) 
```

The above result set is empty so none of the unemployed borrowers had low credit
scores.

Lets analyze the type of loan (ListingCategory) taken by unemployed folks

```{r echo=FALSE, warning = FALSE}
ggplot(data = loan.unemp, aes(x=ListingCategory))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = loan.unemp, aes(x=ListingCategory))+
  geom_bar(aes(y=100*(..count..)/sum(..count..)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Based on the above graph appox. 33% (270) of the total unemployed (835) have 
'Debt Consolidation' as their ListingCategory.

18% have specified 'other',12% have not provided this info, 11% have listed 
'Business', 7% have listed 'household expenses' and all other Listing categories
are below 5% each.

```{r echo=FALSE, warning = FALSE}
#including only numrecic columns for correlation matrix
loan.num_df <- subset(loan, 
                      select = c('Term','DelinquenciesLast7Years',
                                    'StatedMonthlyIncome','DebtToIncomeRatio',
                                    'LoanOriginalAmount','CreditScoreMean'))

loan.corrp_mat <- rcorr(as.matrix(loan.num_df))
loan.corr_mat<- cor(loan.num_df, method = "pearson", use = "complete.obs")

```

Taking help from 
http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-
analyze-format-and-visualize-a-correlation-matrix-using-r-software
and http://jamesmarquezportfolio.com/correlation_matrices_in_r.html to visualize
the correlation matrix.

```{r echo=FALSE, warning = FALSE}
#psych package
pairs.panels(loan.num_df, scale=TRUE)


#corrplot package
corrplot(loan.corrp_mat$r, type="upper", order="hclust", tl.col = "black",
         p.mat = loan.corrp_mat$P, sig.level = 0.01, insig = "blank")

```

The above plot displays positive correlation in blue and negative correlations 
in red. The size of the circle displays the strength of the relation and the
blank cells indicate insignificant relations. The significance level used is 
0.01
The only meaningful relations are between loan original amount and 
creditScoreMean; and loan original amount and term.

##Summarizing findings of Bivariate Analysis
Loan Amount vs. CreditScoreMean yeilds a left skewed distribution. 
With an increase in the creditScoreMean borrowers are able to avail
higher amount loans as expected. The surprising aspect of the plot is loans 
lent to borrowers with low credit scores (almost close to 10).

The correlation coefficient is 0.3408, which means the relation is positive and meaningful but not very strong.

Percentage of defaulters/chargedoff is much higher with low credit
score borrowers.

The lower the Debt-to-Income ratio, the higher are the delinquencies 
(coef = - 0.04387671). So the correlation is negative but not meaningful as it's absolute value < 0.3.

Borrowers with large number of past delinquencies have taken smaller loan 
amounts. (coeff = -0.1366667.)
So the correlation is negative between loan amount and delinquencies but not meaningful as it's absolute value < 0.3

Borrowers with one or more past deliquencies have higher percentages of 
chargedoff/defaulting loans vs zero deliquencies.

Comparing employed and unemployed borrowers, the percentage of loans getting 
chargedoff is very high with unemployed borrowers.

50% of unemployed borrowers availed loan amounts between $2500 and $6000. 
Median being $4000 and maximum being $25000. 

Out of 113937 borrowers, 699 are students. It was fascinating to find only about 3% students unemployed. Rest were either 
working part-time, full-time, self-employed, etc.

It is interesting to see that around 51% borrowers have borrowed money to pay 
off their other loans (based on ListingCategory : Debt Consolidation)
 
Based on their LoanStatus approx. 22% of debt consolidators have completed
their loan, 6% chargedoff, 2% defaulted, whereas about 67% have status listed as
'current' which makes me think if all Debt Consolidation related loans are 
pretty recent compared to other loans.

None of the unemployed borrowers had low credit scores.

Based on the above graph appox. 33% (270) of the total unemployed (835) have 
'Debt Consolidation' as their ListingCategory.
18% have specified 'other',12% have not provided this info, 11% have listed 
'Business', 7% have listed 'household expenses' and all other Listing categories
are below 5% each.

The only meaningful relations are between loan original amount and 
creditScoreMean; and loan original amount and term.


# Multivariate Analysis

```{r echo=FALSE, warning = FALSE,Multivariate_Plots}
ggplot(data = loan, aes(x = CreditScoreMean, y = LoanOriginalAmount))+
  geom_point(alpha = 0.1)+
  facet_wrap(~Term)
```

Let's see the same plots above using boxplots and removing creditScores < 10 to 
correctly see the creditScore mean spread.


```{r echo=FALSE, warning = FALSE}
ggplot(data = loan.high_credit,aes(x = "",y = LoanOriginalAmount))+
  geom_boxplot()+
  facet_wrap(~Term)

ggplot(data = loan.high_credit,aes(x = "",y = CreditScoreMean))+
  geom_boxplot()+
  facet_wrap(~Term)
```

We see that borrowers with higher credit scores (between 600 and 900) have terms
of 12 months and 60 months, while lower credit borrowers have term of 36 months.
Also with increase in term the loan amount is also increasing.

Let's see if we can incorporate loan status in our above geom_point plot.
Also I wish to focus only 3 levels of loan status viz. Completed, defaulted and
chargedoff.
Subsetting loan dataframe to include rows with only "Chargedoff","Completed" 
and "Defaulted" cases. We will look at only complete cases.

```{r echo=FALSE, warning = FALSE}
loan.ls<-subset(loan,LoanStatus %in% c("Chargedoff","Completed","Defaulted"))
loan.ls$LoanStatus<-factor(loan.ls$LoanStatus,
                  levels=c('Completed','Defaulted','Chargedoff'),
                  ordered=FALSE)

loan.ls<-loan.ls[complete.cases(loan.ls), ] 
loan.ls<-na.omit(loan.ls)

```

```{r echo=FALSE, warning = FALSE}

ggplot(data = loan.ls,aes(x = CreditScoreMean,y = LoanOriginalAmount))+
  geom_boxplot(aes(color = LoanStatus, alpha = 0.3))+
  facet_wrap(~Term)+
  scale_color_manual(values=c("orange", "blue", "red"))+
  scale_x_continuous(breaks = seq(0, 1000, 200))


```

From the plot above we see that:
Defaulting/ Charged-off loan amounts < = $25,000.
The defaulters with 36-months term have credit scores in range 50-430.
The chargedoff loans with 36-months term have credit scores in range 440-830. 
Completion of loan is 100% for loan amounts > $25,000

NOTE:
PLEASE COMMENT ON THE CORRECTNESS OF THE MATHEMATICAL MODEL BELOW.
IN THE PREVIOUS REVIEW I RECEIVED NO FEEDBACK FOR THIS SECTION.
THANKS!!!

Now let us try to predict CreditScoreMean based on these variables using a
linear model.

```{r echo=FALSE, warning = FALSE}
m1<-lm(CreditScoreMean ~ LoanOriginalAmount, data = loan)
m2<-update(m1, ~. + DelinquenciesLast7Years )
m3<-update(m2, ~. + StatedMonthlyIncome )
m4<-update(m3, ~. + Term)

mtable(m1,m2,m3,m4)
```

Looking at the low R-squared values we see that the data is heavily scattered
around the linear model described by above code with model m1 being the worst of
all four models. Thus the above models are not good if we are looking for 
precise CreditScore predictions.

Let's see if we can predict LoanStatus based on the other variables. I will only
try to predict if a person can complete, default or get chargedoff on his loan.

Now taking help from
https://datascienceplus.com/perform-logistic-regression-in-r/, I will
apply the Multinom logistic regression model to predict Loan Status.
As my outcome (LoanStatus) variable is categorical (not ordinal) and has more 
than two levels, this model seems appropriate.

```{r echo=FALSE, warning = FALSE}
loan.ls$LoanStatus <- relevel(loan.ls$LoanStatus, ref = "Completed")

```

Now let us split the loan.ls dataset into training and testing subsets for 
modeling (using 75% data) and predicting (using 25% data) .

```{r echo=FALSE, warning = FALSE}
loan.ls.train <- loan.ls[1:37400,] #approximately 75% data
loan.ls.test <- loan.ls[37401:49979,] #approximately 25% data
```


```{r echo=FALSE, warning = FALSE}
# loan.ls1<-loan.ls
# numcols <- grep("^c\\.",names(loan.ls1))
# loan.ls2 <- loan.ls1
# loan.ls2[,numcols] <- scale(loan.ls2[,numcols])
# #m1_sc <- update(m1,data=loan.ls2)
# 
# 
# model.clm <- clm(LoanStatus ~ CreditScoreMean + LoanOriginalAmount +
#             DelinquenciesLast7Years + StatedMonthlyIncome, data = loan.ls1)
# model.clm_sc <- update(model.clm,data=loan.ls2)
# 
# summary(model.clm)
# summary(model.clm_sc)
```


```{r echo=FALSE, warning = FALSE}

m1<- multinom(LoanStatus ~ 1, data = loan.ls.train)
m2<- update(m1, ~. + CreditScoreMean)
m3<- update(m2, ~. + LoanOriginalAmount)
m4<- update(m3, ~. + DelinquenciesLast7Years)
m5<- update(m4, ~. + StatedMonthlyIncome)

mtable(m1, m2, m3, m4,m5, sdigits = 3)

# z <- m5$coefficients/m5$standard.errors
# # 2-tailed z test
# p <- (1 - pnorm(abs(z), 0, 1)) * 2
# exp(coef(m5))

```

Comparing the m1, m2,m3,m4 and m5 models we see that model m5 has lowest 
deviance and AIC.
Now let's apply the Multinomial logistic regression m5 model to fit the training 
data and then predict (correctly classify the loan status) the loan status using
the test dataset 

```{r echo=FALSE, warning = FALSE}
fit.multinom <- multinom(LoanStatus ~ CreditScoreMean + LoanOriginalAmount +
                        DelinquenciesLast7Years + StatedMonthlyIncome,
                        data = loan.ls.train)
 

```

```{r echo=FALSE, warning = FALSE}

pred.multinom <- predict(fit.multinom,newdata=loan.ls.test,type='class')
pred<-as.vector(pred.multinom)
original<-as.vector(loan.ls.test$LoanStatus)

t<-ifelse(pred==original,1,0)

accuracy = sum(t>0)/length(t)
```
The calculated accuracy for correct classification of Loan Status for the model
created is 69.401%. As the Loan Status has 3 levels, the minimum accuracy
expected is 33.33%. Thus an accuracy of 69.4% is pretty decent.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning = FALSE, Plot_One}
ggplot(data = loan, aes(x = CreditScoreMean, y = LoanOriginalAmount))+
  geom_point(alpha = 0.1)+ggtitle("Loan Amount vs. Credit Score")
```

### Loan Amount Vs. Credit Score (Mean)
This plot gives us a good idea of how the loan amounts vary based on Credit 
Scores. The higher credit scores are availing higher loan amounts, but some 
borrowers with almost 10 credit score were also able to avail loans; may be they
had not built a credit history yet as they may be immigrants new to the country, 
or never used credit card before, etc.


### Plot Two
```{r echo=FALSE, warning = FALSE,Plot_Two}
ggplot(data = loan.ls, aes(x = LoanStatus, y = LoanOriginalAmount))+
  geom_boxplot()+
  ggtitle("Loan Amount vs. Loan Status")

```

### Loan Original Amount vs. Loan Status
In general we see higher completion rate for higher loan amounts, the defaulting
loans as well as Chargedoff loans are not more than $25000.

### Plot Three
```{r echo=FALSE,warning = FALSE, Plot_Three}
ggplot(data = loan.ls,aes(x = CreditScoreMean,y = LoanOriginalAmount))+
  geom_point(aes(color = LoanStatus, alpha = 0.3))+
  facet_wrap(~Term)+
  scale_color_manual(values=c("orange", "blue", "red"))+
  scale_x_continuous(breaks = seq(0, 1000, 200))


```

### Loan Amount Vs. Credit Score (Mean) facetted by Term, colored by LoanStatus
This plot brings all the key findings together. 
For 12 months term the variability is the lowest for credit score means and 
LoanAmounts too. The defaulting and chargedoff loans are least for 12 months 
term.

For 36 months term the borrowers with chargedoff loans are surprisingly in the 
higher credit score bracket 450-850 approx. The credit score variability is 
highest for 36 months term.

For 60 months term the spread is highest for Loan Amounts, while credit scores 
lie between 600 and 850.


------

# Reflection
I went through a bit of struggle when trying to predict categorical variable
'LoanStatus' using quantitative independent variables.

During all my exploration I got very few meaningful and zero strong relations 
between the variables chosen. This was surprising for me since I chose these
variables thinking that they are the key factors that can determine loan payment
status for a borrower.

I believe going forward I would like to explore a few other variables to see if 
they are better predictors of loan payment status.
